---
title: "Statistical inference with the GSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data



```{r load-data}
load("gss.Rdata")
```



* * *

## Part 1: Data

According to the survey documentation,
there are different data collection methods, used throughout many years.
Stratification and selection methods are different in different years.
Thus, it is not a 'classical simple random sampling'.

They tell:

"In general, the GSS samples closely resemble distributions reported in the Census and other authoritative sources. Because of
survey non-response, sampling variation, and various other factors the GSS sample does deviate from known population figures for
some variables. The GSS does not calculate any post-stratification weights to adjust for such differences."

## My conclusion:

Generalization: Limited , considering selection methods, bias and under/over represented categories in different years 

Causality: Impossible, while no experiment with the control group has been conducted 

Note:

For the purpose of this test we should assume that the sampling data does indeed reliably represents a certain population. 


* * *

## Part 2: Research question
The research will analyze the possible dependency between major party affiliation and high-income.
Is it true that some parties attract more high-income (subjectively rich) people than other parties in USA?

Spoiler: yes, it seems to be true and I really happy to find this out...
* * *

## Part 3: Exploratory data analysis


Observe how many observations collected per year to get a sense of the size of the sample

```{r}
unique(gss$year)

gss%>%
  group_by(year)%>%
  summarise(people=n())
```

<br><b>
Interpretation:
Looks like each year represented by the significant number of samples</b>
<br>

### Note: This research will set a variable <b>high_income_mark</b>=$60K
### People earning 60K and more will be considered "high income" or "rich"


### I have to check that S-F conditions will be satisfied for my particular research

<b>
Inspecting S-F condition for proportion relative analysis (considering income/party)
This code will fail the entire operation, if the condition (S,F >= 10) fails
</b>


```{r}
high_income_mark<-60000

republicans<-gss%>%
  group_by(year)%>%
  filter((partyid == "Strong Republican" |  partyid == "Not Str Republican") & !is.na(coninc) )%>%
  select(year, partyid, coninc) %>%
  summarise(low_income=sum(coninc < high_income_mark), 
            high_income=sum(coninc >= high_income_mark))

others<-gss%>%
  group_by(year)%>%
  filter((partyid != "Strong Republican" &  partyid != "Not Str Republican") & !is.na(coninc) )%>%
  select(year, partyid, coninc) %>%
  summarise(low_income=sum(coninc < high_income_mark), 
            high_income=sum(coninc >= high_income_mark))

#warn and stop if S-F check fails
if(!all(others$low_income >= 10, others$high_income >= 10, republicans$low_income >= 10, republicans$high_income >= 10)){
  warning("S-F condition has not been met!")
  stop()
}

democrats<-gss%>%
  group_by(year)%>%
  filter((partyid == "Strong Democrat" |  partyid == "Not Str Democrat") & !is.na(coninc) )%>%
  select(year, partyid, coninc) %>%
  summarise(low_income=sum(coninc < high_income_mark), 
            high_income=sum(coninc >= high_income_mark))

others<-gss%>%
  group_by(year)%>%
  filter((partyid != "Strong Democrat" &  partyid != "Not Str Democrat") & !is.na(coninc) )%>%
  select(year, partyid, coninc) %>%
  summarise(low_income=sum(coninc < high_income_mark), 
            high_income=sum(coninc >= high_income_mark))

#warn and stop if S-F check fails
if(!all(others$low_income >= 10, others$high_income >= 10, democrats$low_income >= 10, democrats$high_income >= 10)){
  warning("S-F condition has not been met!")
  stop()
}

democrats


```



### High income Democrats proportion by year

```{r}
#Convert democrats income level to proportion from the total number of democrats
democrats<-gss%>%
  filter((partyid == "Strong Democrat" |  partyid == "Not Str Democrat") & !is.na(coninc) )%>%
  mutate(high_income=ifelse(coninc >= high_income_mark,"high","low"))

democrats_prop<-democrats%>%
  group_by(year)%>%
  summarise(high_inc_prop=sum(high_income=="high")/n())

#Display democrats income level proportions by year
ggplot(democrats_prop, aes(x=year, y = value, color = variable)) +
    geom_line(aes(y = high_inc_prop, col = "high_inc_prop"))+
    geom_point(aes(y = high_inc_prop, col = "high_inc_prop"))+
    labs(title="Democrats proportion with high income by year")
```


<br><b>
Interpretation for the sample:
Proportion of democrats with high income grows</b>
<br>


```{r}
#Convert democrats income level to proportion from the total number of democrats
republicans<-gss%>%
  filter((partyid == "Strong Republican" |  partyid == "Not Str Republican") & !is.na(coninc) )%>%
  mutate(high_income=ifelse(coninc >= high_income_mark,"high","low"))

republicans_prop<-republicans%>%
  group_by(year)%>%
  summarise(high_inc_prop=sum(high_income=="high")/n())

#Display democrats income level proportions by year
ggplot(republicans_prop, aes(x=year, y = value, color = variable)) +
    geom_line(aes(y = high_inc_prop, col = "high_inc_prop"))+
    geom_point(aes(y = high_inc_prop, col = "high_inc_prop"))+
    labs(title="Republicans proportion with high income by year")
```
<br>
<b>
Interpretation for the sample:
Proportion of republicans with high income grows</b>
<br>


### Lets see where is more pronounced high-income voters proportion  
```{r}
both_parties_income_rates<-merge(democrats_prop, republicans_prop, by = c("year", "year"),suffixes = c(".dem",".rep"))
both_parties_income_rates

#lets see all the ratio together while we 
ggplot(both_parties_income_rates, aes(x=year, y = value, color = variable)) +
    geom_line(aes(y = high_inc_prop.dem, col = "high_inc_prop.dem"))+
    geom_point(aes(y = high_inc_prop.dem, col = "high_inc_prop.dem"))+
    geom_line(aes(y = high_inc_prop.rep, col = "high_inc_prop.rep"))+
    geom_point(aes(y = high_inc_prop.rep, col = "high_inc_prop.rep"))+
    scale_y_continuous(breaks = round(seq(0, 0.8, by = 0.05),1))+
    labs(title="Both parties income proportions by year")
```
<br>
<b>
Interpretation for the sample:
Proportion of republicans with high income is always higher throughout all the years
</b>
<br>



* * *

## Part 4: Inference

Note, there is no CI for Chi Square independence test
<br>
<b>So, this research will only include the hypothesis and Chi Square independence test</b>

### Hypothesis Test (Chi-Squared independence test)

<b>
<br>
H_0: There is no dependency between supporting one of two major parties and being rich (60k+ income) 
<br>
H_A: There is a dependency between supporting one of two major parties and being rich  (60k+ income) 
<br>
</b>

Explanatory variable: party
<br>
Response variable:    income
<br>

Note: Though it is presumably a 'strange' dependency: 'particular party affiliation influences income',
considering the absence of the experiment and a control group, 
<b>no one can claim that there is ANY causality at all</b>,
so, this direction of dependency has the same theoretical validity as an opposite direction:
'income influences particular party affiliation'

One may claim, that the same rich party members meet at the same golf clubs 
and thus do create business connections and other conditions for income growth.

So,put aside the 'chicken and eggs' issue, in this Hypothesis we only test
<b>the general presence of dependency (rich<->party affiliation) and causality is out of the scope</b>.


### Lets create unctions to compose chi-square tables for both parties

```{r}


filter_na_income_party <- function(year_param){
  partisans_income<-gss%>%
    filter(!is.na(coninc) & !is.na(partyid) & year==year_param)
  partisans_income
}

get_chi_table_income<- function(partisans_income, party){
    partisans_income<-partisans_income%>%
    mutate(high_income=ifelse(coninc >= high_income_mark, "high","na"))
    proportions_partisan<-partisans_income%>%
      summarise(rich_partisan=sum(being_partisan==party & high_income=="high"),
                rich_other=sum(being_partisan=="other" & high_income=="high"),
                total_rich=sum(high_income=="high"),
                not_rich_partisan=sum(being_partisan==party & high_income=="na"),
                not_rich_other=sum(being_partisan=="other" & high_income=="na"),
                total_not_rich=sum(high_income=="na"),
                total_rep=sum(being_partisan==party),
                total_other=sum(being_partisan=="other"),
                total_all=n())
    
    chi_table <- matrix(c(proportions_partisan$rich_partisan, proportions_partisan$rich_other,proportions_partisan$total_rich, 
                    proportions_partisan$not_rich_partisan, proportions_partisan$not_rich_other, proportions_partisan$total_not_rich,
                    proportions_partisan$total_rep,proportions_partisan$total_other,proportions_partisan$total_all), ncol=3, byrow=TRUE)
    colnames(chi_table) <- c(party,'other','total')
    rownames(chi_table) <- c('rich','not_rich', 'total')
    chi_table <- as.table(chi_table)
    chi_table
}



get_chi_table_income_republicans <- function(year_param) {
  partisans_income<-filter_na_income_party(year_param)
  partisans_income<-partisans_income%>%
    mutate(being_partisan=ifelse(partyid == "Strong Republican" |  partyid == "Not Str Republican","republican","other"))
  get_chi_table_income(partisans_income,"republican")
}

get_chi_table_income_democrats <- function(year_param) {
  partisans_income<-filter_na_income_party(year_param)
  partisans_income<-partisans_income%>%
    mutate(being_partisan=ifelse(partyid == "Strong Democrat" |  partyid == "Not Str Democrat","democrat","other"))
  get_chi_table_income(partisans_income,"democrat")
}


```

### Here you can see example Chi-Square tables for each party for 1 selected year

```{r}
#Example(chi-table construction by year):
rep_income<-get_chi_table_income_republicans("1972")
rep_income

dem_income<-get_chi_table_income_democrats("1972")
dem_income

```

### Lets create a function for p-value calculation 


Note, returned is not only p-value, but a direction of discovered dependency:
<br>
If, for example, there is a positive correlation between party 
affiliation and increased proportion of rich, it is direction=2.
<br>
The negative correlation produces direction 1.
<br>
Those directions we later need to construct visual presentation.
<br>

Note: This function also validates Chi-table population parameters
<br>
```{r}
getDependencyPValue <- function(party_income) {
  
  totalRichProp    <- party_income[1,3]/party_income[3,3]
  totalNotRichProp <- 1-totalRichProp
  
  
  pop_rich_partisan        <- party_income[3,1]*totalRichProp
  pop_not_reach_partisan   <- party_income[3,1]*totalNotRichProp
  pop_rich_other           <- party_income[3,2]*totalRichProp
  pop_not_rich_others      <- party_income[3,2]*totalNotRichProp
  
  
  tol = 1e-5# tolerance level for floats comparison
  
  ##Lets check that calculated population numbers are fit
  if(abs(pop_rich_partisan+pop_not_reach_partisan-party_income[3,1]) > tol){
    warning("Failed to conctruct expected population of partisans")
    stop()
  }
  if(abs(pop_rich_other+pop_not_rich_others-party_income[3,2]) > tol){
    warning("Failed to conctruct expected population of others")
    stop()
  }
  if(abs(pop_rich_partisan+pop_rich_other-party_income[1,3]) > tol){
    warning("Failed to conctruct expected population of rich")
    stop()
  }
  if(abs(pop_not_reach_partisan+pop_not_rich_others-party_income[2,3]) > tol){
    warning("Failed to conctruct expected population of not rich")
    stop()
  }
  
  Chi2<-((party_income[1,1]-pop_rich_partisan)**2/pop_rich_partisan+
       (party_income[2,1]-pop_not_reach_partisan)**2/pop_not_reach_partisan+ 
       (party_income[1,2]-pop_rich_other)**2/pop_rich_other+
       (party_income[2,2]-pop_not_rich_others)**2/pop_not_rich_others)  

  DF=(2-1) * (2-1) # We always have 2 columns, 2 rows
  p_value=pchisq(Chi2, DF, lower.tail = FALSE)
  
  #the direction of the disproportion in rich partisans in observed data
  direction<-0 # by default there is no disproportion
  if(party_income[1,1]/party_income[3,1]<party_income[1,2]/party_income[3,2]){
      direction<-1 #partisans are less rich as all others
  }else if(party_income[1,1]/party_income[3,1]>party_income[1,2]/party_income[3,2]){
      direction<-2 #partisans are more rich as all others
  }
  
  annual_party_result <- list("pvalue" = p_value, "direction" = direction)
  return(annual_party_result) 
  
}

```


### Example run of the function for 1 particular party for 1 particular year

```{r}
party_income<-get_chi_table_income_republicans("2012")
annual_party_result = getDependencyPValue(party_income)
pvalue=annual_party_result$pvalue
direction=annual_party_result$direction
pvalue
direction

```
<br>
<b>
Interpretation:
For this year: 2012 p_value is << 0.05(H_o rejected), 
direction is 2, republicans do have visibly (statistically) increased proportion of rich among their voters
</b>


### Lets prepare functions to run thus HT for all years for two parties

```{r}

years<-sort(unique(gss$year))
getPartyInfo <- function(years, party) {
  p_values<-vector("numeric", length(years))
  directions<-vector("numeric", length(years))
  index=0
  for(year in years) {  
      party_income
      if(party=="democrats"){
         party_income<-get_chi_table_income_democrats(year)
      }else if(party=="republicans"){
         party_income<-get_chi_table_income_republicans(year)
      }
      annual_party_result = getDependencyPValue(party_income)
      pvalue=annual_party_result$pvalue
      direction=annual_party_result$direction
      p_values[index]<-pvalue
      directions[index]<-direction
      index <- index+1
  }
  party.info <- data.frame(years, p_values, directions)
  party.info
}

dems.info = getPartyInfo(years,"democrats")
reps.info = getPartyInfo(years,"republicans")

```

<br>
<b>
Interpretation: two summary tables above show p-value and direction for two parties</b>


### Lets merge two parties together

```{r}
both_parties_chi_data<-merge(dems.info, reps.info, by = c("years", "years"),suffixes = c(".dem",".rep"))
both_parties_chi_data
```


### Lets normalize values

<b>
1. p-values>0.05 converted to 0: no significance (no dependency found)
<br>
2. p-values go through the log scaling to better show them in the graph
<br>
3. Dependent on direction p-values changed to negative (direction=1) to show it better in the graph
(In reality Chi Square distribution has no negative values, we know that!)
<br>
</b>

```{r}

both_parties_chi_data<-both_parties_chi_data%>% 
  mutate(norm_dem_pvalue=ifelse(p_values.dem<0.05,log(p_values.dem),0))%>% 
  mutate(norm_rep_pvalue=ifelse(p_values.rep<0.05,log(p_values.rep),0))

both_parties_chi_data<-both_parties_chi_data%>% 
  mutate(signed_norm_dem_pvalue=ifelse(directions.dem==2,-norm_dem_pvalue,ifelse(directions.dem==1,norm_dem_pvalue,0)))%>% 
  mutate(signed_norm_rep_pvalue=ifelse(directions.rep==2,-norm_rep_pvalue,ifelse(directions.rep==1,norm_rep_pvalue,0)))

both_parties_chi_data
```



```{r}
ggplot(both_parties_chi_data, aes(x=years, y = value, color = variable)) +
    geom_line(aes(y = signed_norm_dem_pvalue, col = "dem_rich_dependency"))+
    geom_point(aes(y = signed_norm_dem_pvalue, col = "dem_rich_dependency"))+
    geom_line(aes(y = signed_norm_rep_pvalue, col = "rep_rich_dependency"))+
    geom_point(aes(y = signed_norm_rep_pvalue, col = "rep_rich_dependency"))+
    scale_x_continuous(breaks = round(seq(years[1], years[length(years)], by = 1),1))+
    labs(title="Both parties high income affiliation dependency")+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

```
<b>
<br>
Interpretation and conclusion:
</b>
<br>
Proportion of rich among republicans is consistently high.(There is a dependency!)
<br>
Proportion of rich among democrats is consistently low (There is a dependency!)
<br>
In 2005 it was very much pronounced (George W. Bush was a President)
<br>
In 2012 both proportions become almost equal.
<br>
So, generally, for those years, we have visible tendency of rich voters to prefer Republicans
<br>
OR
<br>
Republicans to accommodate increased number of rich.
<br>
Democrats do have invert proportion of rich accommodation.
<br>

But: According to surveys published in very recent years, this tendency seems to be minimized now.
<br>
https://www.cnbc.com/2016/10/23/the-rich-vote-republican-maybe-not-this-time.html
<br>
(Unfortunately we do not have latest years in the data to prove that)






